A Study of Common Bug Fix Patterns in Rust \thanks{Competing Interests: We have no competing interests and are funded by a Discovery Grant from Canada's Natural Science and Engineering Research Council.}

### 2023-06-05 discussion.tex
reread for usefulness

In this work, we used Ruxanne to successfully mine 20 cross-project bug fix patterns. We presented these patterns in two groups: 12 general patterns and 8 BC-related patterns. Clippy is able to detect 5 of the 8 BC-related patterns. Fundamentally, Clippy cannot detect some of the remaining patterns. For instance, we proposed pattern \textit{Adding mutability}. In the Rust context, this change anticipates a future change to the code which will perform a mutation of a previously immutable value. Predicting the future is beyond Clippy's remit. To the best of our knowledge, the remaining patterns (3 BC-related patterns and 12 general patterns) cannot be reported by the current linting tools.

@@

When presenting each pattern, we discussed actionability for the most important actions. Here, we discuss the implications of our most important patterns. In summary, program repair tools must understand semantics of attributes and should be performance-aware. It may also be useful for tools to be able to automatically infer immutability.

The two most common patterns were general patterns modifying struct fields and attributes, with 179 and 172 instances respectively. Unfortunately, struct field modifications are neither interesting nor actionable; they are simply used to enable other changes. Attribute modifications are far more interesting, and account for the second- and fifth-most popular patterns. From a language design perspective, the fact that there are so many attribute changes marked as bug fixes illustrates the semantic importance of attributes in Rust, and \emph{implies that program repair must understand the effects of attributes}. Yet, different attributes have different, and often library-dependent, semantic meanings. Proposing specific repair techniques that reason about attributes is beyond the scope of this work, but our results suggest that they are important to develop.

@@

@@

For instance, as we have discussed, for the common pattern of \textit{Modifying the attributes of structs}, a code linter tool could search for correct attribute lists within the codebase~\citep{forrest2009genetic}, leveraging the assumption that proper attribute usage can be found elsewhere. Also, the linter can analyze the usages of the struct to determine the necessity of adding the attribute. Removing unnecessary attributes not only optimizes the codebase but also eliminates unnecessary overhead and reduces binary size.

>>>>

In this work, we used Ruxanne to successfully mine 20 cross-project bug fix patterns. We presented these patterns in two groups: 12 general patterns and 8 BC-related patterns. Clippy is able to detect 5 of the 8 BC-related patterns. Fundamentally, Clippy cannot detect some of the remaining patterns. For instance, we proposed pattern \textit{Adding mutability}. In the Rust context, this change anticipates a future change to the code which will perform a mutation of a previously immutable value. Predicting the future is beyond Clippy's remit. To our knowledge, the remaining patterns (3 BC-related patterns and 12 general patterns) cannot be reported by current linting tools.

@@

When presenting each pattern, we discussed actionability for the most important actions. Here, we discuss the implications of our most important patterns. \emph{In summary, program repair tools must understand semantics of attributes and should be performance-aware.} It may also be useful for tools to be able to automatically infer immutability.

\todo{is the most common ordering still the same after trimming?}
The two most common patterns were general patterns modifying struct fields and attributes, with 179 and 172 \todo{are those numbers stale?} instances respectively. Unfortunately, struct field modifications are neither interesting nor actionable; they are simply used to enable other changes. Attribute modifications are far more interesting, and account for the second- and fifth-most popular patterns. From a language design perspective, the fact that there are so many attribute changes marked as bug fixes illustrates the semantic importance of attributes in Rust, and \emph{implies that program repair must understand the effects of attributes}. Yet, different attributes have different, and often library-dependent, semantic meanings. Proposing specific repair techniques that reason about attributes is beyond the scope of this work, but our results suggest that they are important to develop.

@@

\todo{recap the italicized stuff from above}
\vspace*{1em}
\fbox{\parbox{0.9\textwidth}{We summarize learnings from our bug patterns as follows:
\begin{itemize}
\item one
\item two
\end{itemize}
}}

@@

For instance, as we have discussed, for the common pattern of \textit{Modifying the attributes of structs}, a code linter tool could search for correct attribute lists within the codebase~\citep{forrest2009genetic}, leveraging the assumption that proper attribute usage can be found elsewhere in the code. Also, the linter can analyze the usages of the struct to determine the necessity of adding the attribute. Removing unnecessary attributes not only optimizes the codebase but also eliminates unnecessary overhead and reduces binary size.

### 2023-06-05 introduction.tex
reread for usefulness

@@

In this work, with the help of our tool, Ruxanne, we empirically analyze the most popular Rust projects in GitHub. Our goal is to discover common bug patterns that afflict these projects. Although we are motivated by a desire to develop automated program repair tools, we believe these insights can also be of broader service. Automated program repair~\citep{goues2019automated,liu2018survey} tools try to find bug locations in the source code (using a fault localization module) and then generate patches to fix them, possibly without human intervention, so that the modified program meets desired specifications (implicit or explicit). A powerful automated program repair system has many useful applications and could significantly simplify debugging, thus reducing software development cost~\citep{le2012systematic}. A set of common bug fix patterns can be a useful input to an APR system: it tells how to fix recurring buggy structures, thus reducing the search space for patch generation~\citep{jeffrey2009bugfix}. 
Yet, to the best of our knowledge, no research has sought to identify the most pervasive bug fix patterns in Rust programs. In this project, we introduce Ruxanne, a tool for analyzing and extracting fix patterns in Rust. Ruxanne uses a novel embedding of Rust code into fixed-sized vectors. Through Ruxanne, we mined the top 18 most-starred Rust projects in Github to discover the most common bug fix patterns committed to their repositories. We analyzed 87,726 code changes drawn from 57,214 commits across these 18 projects. After clustering the code changes, and conducting a manual analysis, we discovered 20 groups of cross-project bug fix patterns, which we categorize as (1) general patterns and (2) borrow-checker-related patterns. We describe each of these patterns and their implications to automated program repair.

If we observe undesired behaviour from our program, then there is likely a fault within our program\footnote{At the extreme, a fault could be caused by moths or other hardware malfunctions.}. But where is the fault? Fault localization~\citep{wong2016survey} is the process of automatically finding faulty statements in a program. If a fault localization tool knows about the common bug patterns that can appear in the underlying programming language, it can prioritize candidates based on their bug-producing potential. That would result in a more robust tool---one that leverages historical data.

Moreover, knowing that certain code patterns in a programming language are more susceptible to bugs enables projects to set policies that avoid such patterns. For instance, in the context of C and C++, \cite{Cannon1997RecommendedCS} recommend that, if at least one of the if/else sections is a compound statement, requiring braces, then both sections should have braces (i.e. they should be fully bracketed). This recommendation presumably comes from qualitative experience, as it pre-dates large-scale empirical studies. A project can choose to require that contributions follow such recommendations. Research such as ours can help formulate data-driven recommendations that reduce the number of patterns empirically linked to bugs. Per Monperrus, the fewer frequent bug patterns we keep in our software, the lower cost we can expect to pay for software maintenance~\citep{monperrus2014critical}.

@@

\begin{itemize}
    \item To the best of our knowledge, this work is the first to exploit a code analysis pipeline to automatically mine Rust open source projects and extract common cross-project bug fix patterns.
    \item This work proposes a novel method to encode key AST information in fixed size datapoints using a semi-automatically derived weighting scheme.
    \item Keeping in mind the key role of the borrow checker in Rust, this work specifically identifies bug fix patterns that are related to Rust's ownership framework in addition to general bug fix patterns for Rust.
\end{itemize}

>>>>

Rust is a relatively new programming language~\citep{klabnik2019rust} which allows programmers to write programs that have low-level control over resources, while still ensuring high-level safety guarantees (for programs written in safe Rust). What distinguishes Rust from other systems programming languages (e.g. C/C++) is its ownership framework, which is an integral part of Rust. The ownership framework keeps track of references to different memory locations, and will free them once they go out of scope. If a program passes the compiler’s checks, it meets several safety properties by construction; e.g. it has no dangling references or double-frees. Such properties had to be manually ensured (or not!) in languages like C/C++. Moreover, not only does ownership remove the need for a garbage collector in the runtime environment, it frees the programmer from the responsibility to ensure certain safety-related properties~\citep{qin2020understanding}. Due to the memory and thread safety guarantees that Rust provides, many software companies have adopted Rust in production. For example, Dropbox decided to migrate to Rust for developing their storage system, since they were not completely satisfied with the features provided by AWS’s S3 or Golang~\citep{dropbox}.

@@

In this work, with the help of our tool, Ruxanne, we empirically analyze the most popular Rust projects in GitHub. Our goal is to discover common bug patterns that afflict these projects. Although we are motivated by a desire to develop automated program repair tools, we believe these insights can also be of broader service. Automated program repair~\citep{goues2019automated,liu2018survey} tools try to find bug locations in the source code (using a fault localization module) and then generate patches to fix them, possibly without human intervention, so that the modified program meets desired specifications (implicit or explicit). A powerful automated program repair system has many useful applications and could significantly simplify debugging, thus reducing software development cost~\citep{le2012systematic}. A set of common bug fix patterns can be a useful input to an APR system: it tells how to fix recurring buggy structures, thus reducing the search space for patch generation~\citep{jeffrey2009bugfix}. 
Yet, to the best of our knowledge, no research has sought to identify the most pervasive bug fix patterns in Rust programs.

In this work, we introduce Ruxanne, a tool for analyzing and extracting fix patterns in Rust. Ruxanne uses a novel embedding of Rust code into fixed-sized vectors. Through Ruxanne, we mined the top 18 most-starred Rust projects in Github to discover the most common bug fix patterns committed to their repositories. We analyzed 87,726 code changes drawn from 57,214 commits across these 18 projects. After clustering the code changes, and conducting a manual analysis, we discovered 20 groups of cross-project bug fix patterns, which we categorize as (1) general patterns and (2) borrow-checker-related patterns. We describe each of these patterns. The most common general pattern is addition or removal of struct fields, while the most common borrow-checker-related bug fix pattern is removing a call to the clone() function. Our patterns can serve as knowledge to be embedded into the design of novel Rust automated program repair tools; we present some specific potential applications of these patterns below.

If we observe undesired behaviour from a program, then there is likely a fault within the program\footnote{At the extreme, a fault could be caused by moths or other hardware malfunctions.}. But where is the fault? Fault localization~\citep{wong2016survey} is the process of automatically finding faulty statements in a program. If a fault localization tool knows about the common bug patterns that can appear in the underlying programming language, it can prioritize candidates based on their bug-producing potential. That would result in a more robust tool---one that leverages historical data.

Moreover, knowing that certain code patterns in a programming language are more susceptible to bugs enables projects to set policies that avoid such patterns. For instance, in the context of C and C++, \cite{Cannon1997RecommendedCS} recommend that, if at least one of the if/else sections is a compound statement, requiring braces, then both sections should have braces (i.e. they should be fully bracketed). This recommendation presumably comes from qualitative experience, as it pre-dates large-scale empirical studies. A project can choose to require that contributions follow such recommendations. Research such as ours can help formulate data-driven recommendations that reduce the number of patterns empirically linked to bugs. Monperrus espouses the claim that the fewer frequent bug patterns we keep in our software, the lower cost we can expect to pay for software maintenance~\citep{monperrus2014critical}.

@@

\begin{itemize}
    \item To the best of our knowledge, this work is the first to exploit a code analysis pipeline to automatically mine Rust open source projects and extract common cross-project bug fix patterns.
    \item This work proposes a novel method to encode key AST information in fixed size datapoints using a semi-automatically derived weighting scheme.
    \item Keeping in mind the key role of the borrow checker in Rust, this work specifically identifies bug fix patterns that are related to Rust's ownership framework in addition to general bug fix patterns for Rust.
    \item This work finds that the most frequently observed general pattern is the addition or removal of struct fields and that the most frequently observed BC-related pattern is the removal of a clone() call. 
\end{itemize}

\noindent \textbf{Data Availability Statement} The datasets generated and analyzed during the current study are available in the Zenodo repository, \url{https://zenodo.org/record/7388618}.

### 2023-06-05 main.tex
grammar
reread for usefulness

Yet, to the best of our knowledge, no research has sought to identify the most pervasive bug fix patterns within Rust programs. In this project, we introduce Ruxanne, a tool for analyzing and extracting fix patterns in Rust. Ruxanne implements a novel embedding of Rust code into fixed-sized vectors. Using Ruxanne, we mined the top 18 most-starred Rust projects in Github to discover the most common bug fix patterns committed to their repositories. We analyzed 87,726 code changes drawn from 57,214 commits across these 18 projects. After clustering the code changes, and conducting a manual analysis, we discovered 20 groups of cross-project bug fix patterns, which we categorize as (1) general patterns and (2) borrow-checker-related patterns. Among the general patterns, the most frequently observed pattern is when the user either adds or removes one of the struct fields. In the case of borrow-checker-related patterns, the most common pattern we encountered is when the user removes a clone() call. We describe all the patterns and their implications to automated program repair.
\keywords{Bug patterns \and Pattern mining \and Bug fix changes \and Rust}
% \PACS{PACS code1 \and PACS code2 \and more}
\subclass{68N01 \and 68N15} %% Computer Science : General and Programming Languages
\end{abstract}

>>>>

Yet, to our knowledge, no research has sought to identify the most pervasive bug fix patterns within Rust programs. In this project, we introduce Ruxanne, a tool for analyzing and extracting fix patterns in Rust. Ruxanne implements a novel embedding of Rust code into fixed-sized vectors. Using Ruxanne, we mined the top 18 most-starred Rust projects in Github to discover the most common bug fix patterns committed to their repositories. We analyzed 87,726 code changes drawn from 57,214 commits across these 18 projects. After clustering the code changes, and conducting a manual analysis, we discovered 20 groups of cross-project bug fix patterns, which we categorize as (1) general patterns and (2) borrow-checker-related patterns. Among the general patterns, the most frequently observed pattern is when the user either adds or removes struct fields. In the case of borrow-checker-related patterns, the most common pattern we encountered is when the user removes a clone() call. We describe all detected patterns and their implications to automated program repair.
\keywords{Bug patterns \and Pattern mining \and Bug fix changes \and Rust}
% \PACS{PACS code1 \and PACS code2 \and more}
\subclass{68N01 \and 68N15} %% Computer Science : General and Programming Languages
\end{abstract}

### 2023-06-05 results.tex
reread for usefulness

\textbf{What are the general fix patterns in Rust, and how often do they apply?} 
We collected general fixes, asked DBSCAN to create clusters, and manually analyzed them. In this subsection, we will discuss each pattern, as well as its actionability---possible usages of the pattern for bug repair. Section~\ref{subsec:patterns} contains a global discussion of our patterns and their value.

Specifically, our dataset includes 60583 datapoints for general fix patterns. We processed our dataset on a server with a Xeon Gold 5120 processor (Q3-2017, 14-core 2.2 to 3.2GHz). The data collection took around 26 hours to populate the general and BC-related databases $D_g$ and $D_b$. We used DBSCAN with parameters $\epsilon=0.0001$ and $Z=5$, resulting in 577 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 12 general patterns, which we describe in Table~\ref{table:general}. As shown in the table, we have separated the patterns based on the underlying program element (Category column). We have given each pattern a specific ID (ID column) and determined the number of instances manifesting that fix pattern (\# Instances column), as well as the number of projects in which the pattern was seen (\# Project column).

Since these patterns are general Rust patterns, analogous patterns can also occur in other programming languages, and we will discuss such analogous patterns when describing each pattern. We will also discuss the actionability of some of the patterns. Moreover, other researchers can use the empirical pattern frequency numbers that we report to build their own program repair tools, prioritizing more-frequently-repaired patterns; they are not limited to our suggested actionability. We next provide a detailed description of each of these patterns.

@@

\subsubsection{Attributes}
The most common bug fix pattern in our corpus modifies attributes. In Rust, attributes enable developers to add metadata to program elements. Some of the attributes---presumably all of the ones involved in bug fixes---affect the semantics of the program; for instance, libraries can generate different code based on the attributes that the developer specifies.

Developers can add attributes to structs, struct fields, enum variants, match expression arms, etc. There are two types of attributes\todo{how important is inner versus outer?}: Inner Attributes and Outer Attributes. Inner Attributes apply to the program element encapsulating them, while Outer Attributes apply to the program element coming after the attribute.

The concept of attributes in Rust is similar to Python's decorators and Java's annotations (though Java annotations may be less likely to have runtime effect; for instance, Java @Override adds additional compile-time constraints). While each language may have its own syntax and specific use cases, the fundamental idea of attaching additional information to code elements is shared across these languages. We describe bug patterns that involve modification in Attributes. In the case of Python, according to the findings of \cite{yang2022mining}, no bug patterns involving Python's decorators were reported in their study. \\

@@

\noindent\textbf{Description:} In this pattern, the developer changes the attribute set of a struct. There are three variants of this pattern: Adding attributes to the struct (G.attr.struct.add); Removing attributes from the struct (G.attr.struct.drop); Changing the content of an existing attribute (G.attr.struct.change). These variants account for 58, 10, and 13 out of 81 datapoints in the cluster, respectively. Thus, adding attributes is the most frequent variant. Alse, these attributes are Outer Attributes as we can find struct-related non terminals within the ASTDiff of the instances of this pattern.

In the specific cases above, the Add change requests that additional boilerplate code be generated to evaluate partial equivalence on \texttt{CounterProps} (new functionality); the Remove change causes debug code to be removed (performance fix); and the Change change requests different deserialization code be generated. \\

@@

\noindent\textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, there are also three subtypes of this change pattern. Out of 34 datapoints in this cluster, 32 involve changes to a field attribute, while one datapoint each corresponds to the addition and removal of a field attribute. Using these attributes, the developers can control different actions they wish to apply on struct fields, in addition to providing meta-information. For instance, in the code snippet provided above, the developers chose to make Serde (a serialization/deserialization tool for Rust) call a different function (\verb+is_empty+ instead of \verb+Option::is_none+) to determine whether to skip serializing the \verb+attributes+ field. Note that \textit{\textbf{G.attr.struct.field}} is not a subset of \textit{\textbf{G.attr.struct}}. \\

\noindent\textbf{Actionability:} Because attributes have semantic effects, program repair tools can effect repairs by adjusting the set of attributes included. Starting from the observation that so many of our bug fixes change attributes, we believe that IDEs need to be able to propose relevant attribute changes, as witnessed in our corpus. The example for \textit{\textbf{G.attr.struct.change}} about changing \verb+Deserialize+ to \verb+ConfigDeserialize+ is specific to that program, but other changes would apply more generally. Also, as for \textit{\textbf{G.attr.struct.drop}}, we demonstrated a case where the developer removed a \verb+Debug+ attribute. A novel type of performance program repair tool working with a run-time profiler could identify unnecessary code and remove attributes that generate it.

@@

Like in other languages, such as C, a struct makes it possible for users to define a custom type that is comprised of different types. The patterns that we describe in this section can also be seen in other languages that use the same concept. \\

@@

\noindent\textbf{Description:} Adding \texttt{pub} to a struct field makes it possible to access that field from external modules. In the instances of this cluster, adding the \texttt{pub} access modifier happened in bug fixing changes, where the developer needed to access a field which had not been marked for public access. On the other hand, public access might be revoked if the developer realizes there are no external accesses to the field, and that such accesses are not desirable, e.g. for encapsulation purposes. \\

\noindent\textbf{Actionability:} The patterns related to access modifiers suggest an opportunity for repair tools to make sure that the accessability of program entities follows the user's specifications, enhancing encapsulation, security, and maintainability. Additionally, the empirical numbers revealing the frequency of struct field modifications in Rust shed light on the significance of these bug fix patterns and emphasize the need for further exploration on how program repair tools can effectively modify code based on changes in struct fields to repair program behavior.

@@

\noindent\textbf{Description:} Rust is a null-safe language, meaning that object references cannot take on null values. If a developer wants a variable to contain the equivalent of null, they would have to use the \texttt{Option} type. \texttt{Option} is essentially an \texttt{enum} with two variants, \texttt{Some}, which indicates that a variable has a value, and \texttt{None}, which indicates the absence of any value. In this change pattern, the developer changes the type of a struct field to an \texttt{Option} of that type, hence enabling them to store no value in that field. 
For instance, in the Rust project swc-project/swc (a fast TypeScript/JavaScript compiler), the developer needed to modify the CSS parser source code to account for empty @media queries\footnote{\scriptsize \url{https://github.com/swc-project/swc/commit/75a14f98b7370226115ee24eec6eb8c802bd4837}}. As shown in the snippet above, they changed type \texttt{MediaQueryList} to \verb+Option<MediaQueryList>+, and modified the other parts of the source code accordingly. \\

\noindent\textbf{Actionability:} When the user's specification requires a variable to be able to assume \texttt{None} values, these patterns indicate the need to modify the program accordingly. A program repair\todo{is this really a repair tool or a refactoring tool? I feel like many of our changes are for refactorings not repairs, despite our PR} tool can leverage these patterns to automatically update the variable's type to \texttt{Option<T>}, ensuring that it can now hold \texttt{None} as a valid value. Furthermore, such a tool should also systematically update other occurrences of that variable throughout the program to account for the change and ensure program correctness.

@@

Rust is a statically-typed language, and like in the other statically-typed languages, some change patterns are associated with changes in types used in different program structures. Here, we introduce four bug fix patterns that are associated with types in Rust:

@@

We collected 27143 datapoints for borrow-checker related fix patterns. We applied DBSCAN with parameters $\epsilon=0.001$ and $Z=5$, resulting in 102 clusters. After manual analysis (Section~\ref{sec:manual_analysis_cluster_selection}), we ended up with 8 borrow-checker related fix patterns. Similar to the previous table, we describe BC-related patterns in Table~\ref{table:bc}. Since the Borrow-Checker is unique to Rust, patterns in this section generally do not have direct counterparts in other languages. We next provide a detailed description of each of these patterns.

@@

\noindent\textbf{Actionability:} Like Clippy, a program repair tool can implement clone removal by reasoning about whether the clone operation was necessary. If the clone is unnecessary, the tool can replace it with a reference (as seen above), thus reducing the overhead of unnecessary copying. \\

@@

\noindent\textbf{Actionability:} This pattern often occurs as part of a larger change, making it challenging to create a generalized repair tool that covers all possible scenarios. Typically, when a reference is passed instead of transferring ownership, the variable is used again after returning from the call. Although creating a comprehensive repair tool for such cases is complex, empirical results underscore the significance and prevalence of this fix, warranting further investigation. \\

\noindent\textit{\textbf{BC.deref.to\_string().add} Adding a dereference operation before calling to\_string()}

@@

\noindent\textbf{Description:} This is a change that in all instances (in our benchmarks) has been proposed by Clippy---commit messages indicated the use of Clippy for this fix. The change applies on an object prior to calling \verb+to_string()+ on that object. In this change, a dereference operator is added to that object. The reason behind the change is that the object is a reference type of \verb+T+, while \verb+T+ directly implements \verb+to_string()+. Adding the dereference makes the compiler use the specialized implementation of \verb+to_string()+ and not go through slower string formatting methods\footnote{\url{https://rust-lang.github.io/rust-clippy/master/index.html\#inefficient\_to\_string}}.

@@

\noindent\textbf{Actionability:} A repair tool can use this pattern to introduce mutability to the target variable and modify other parts of the program accordingly. For instance, if the target variable has associated \texttt{impl} blocks, the change may involve modifying method annotations to ensure that the calling object is mutable, aligning with the desired specifications.\todo{how would the specs be provided? how could this be done as part of repair vs as part of refactoring?!} \\

@@

\noindent\textbf{Description:} The developer drops the mut keyword before a variable, as they do not mutate the variable. This redundant mutability is reported by Clippy and can also be removed by it. This change may help future-proof the code against unintended future changes.

>>>>

\textbf{What are the general fix patterns in Rust, and how often do they apply?} 
We collected general fixes, asked DBSCAN to create clusters, and manually analyzed them. In this subsection, we will discuss each pattern, as well as one aspect of its value---its actionability. By actionability, we mean the possible usages of the pattern for bug repair. Other researchers can also use the empirical pattern frequency numbers that we report to build their own program repair tools, prioritizing more-frequently-repaired patterns; they are not limited to our suggested actionability.  Section~\ref{subsec:patterns} contains a global discussion of our patterns and their value.

Our dataset includes 60583 datapoints for general fix patterns. We processed our dataset on a server with a Xeon Gold 5120 processor (Q3-2017, 14-core 2.2 to 3.2GHz). The data collection took around 26 hours to populate the general and BC-related databases $D_g$ and $D_b$. We used DBSCAN with parameters $\epsilon=0.0001$ and $Z=5$, resulting in 577 clusters. After manual analysis, as described in Section~\ref{sec:manual_analysis_cluster_selection}, we ended up with 12 general patterns, which we describe in Table~\ref{table:general}. As shown in the table, we have separated the patterns based on the underlying program element (Category column). We have given each pattern a specific ID (ID column) and determined the number of instances manifesting that fix pattern (\# Instances column), as well as the number of projects in which the pattern was seen (\# Project column).

Since these patterns are general Rust patterns, analogous patterns can also occur in other programming languages, and we will discuss such analogous patterns when describing each pattern. We next provide a detailed description of each of these patterns.

@@

\subsubsection{Attributes}
The most common bug fix pattern in our corpus modifies attributes. In Rust, attributes enable developers to add metadata to program elements. Some of the attributes---presumably the vast majority of those involved in bug fixes---affect the semantics of the program; for instance, libraries generate different code based on the attributes that the developer specifies.

Developers can add attributes to structs, struct fields, enum variants, match expression arms, etc.
%I don't think we're well served to talk about inner vs outer attributes, not relevant to readers
%There are two types of attributes\todo{how important is inner versus outer?}: Inner Attributes and Outer Attributes. Inner Attributes apply to the program element encapsulating them, while Outer Attributes apply to the program element coming after the attribute.
The concept of attributes in Rust is similar to Python's decorators and Java's annotations. While each language may have its own syntax and specific use cases, the fundamental idea of attaching metadata to code elements is shared across these languages. Some attributes, decorators, and annotations have no run-time effect and thus could not cause bugs (if we accept a purely operational definition of ``bug''). \cite{yang2022mining}~reported no bug patterns involving Python's decorators in their Python dataset. As we alluded to above, however, Rust has many attributes that definitely have run-time effects. We will see some of these attributes in the following examples.

@@

\noindent\textbf{Description:} In this pattern, the developer changes the attribute set of a struct. There are three variants of this pattern: Adding attributes to the struct (G.attr.struct.add); Removing attributes from the struct (G.attr.struct.drop); Changing the content of an existing attribute (G.attr.struct.change). These variants account for 58, 10, and 13 out of 81 datapoints in the cluster, respectively. Thus, adding attributes is the most frequent variant. 
%Also, these attributes are Outer Attributes as we can find struct-related non terminals within the ASTDiff of the instances of this pattern.

In the specific cases above, the Add change requests that additional boilerplate code be generated to evaluate partial equivalence on \texttt{CounterProps} (new functionality); the Remove change causes debug code to be removed (performance fix); and the Change change requests different deserialization code be generated. In our example, adding an attribute causes additional code to be generated for a struct and could thus, for example, fix a compile-time error in a dependency. \\

@@

\noindent\textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, there are also three subtypes of this change pattern. Out of 34 datapoints in this cluster, 32 involve changes to a field attribute, while two datapoints correspond to the addition and removal of a field attribute (one for addition, one for removal). Using these attributes, the developers can control different actions they wish to apply on struct fields, in addition to providing meta-information. For instance, in the code snippet provided above, the developers chose to make Serde (a serialization/deserialization tool for Rust) call a different function (\verb+is_empty+ instead of \verb+Option::is_none+) to determine whether to skip serializing the \verb+attributes+ field. Note that \textit{\textbf{G.attr.struct.field}} is not a subset of \textit{\textbf{G.attr.struct}}. \\

\noindent\textbf{Actionability of attribute change bug fixes:} Because attributes have semantic effects, program repair tools can effect repairs by adjusting the set of attributes included. Given the observation that so many of our bug fixes change attributes, we propose that IDEs need to be able to propose relevant attribute changes. The example for \textit{\textbf{G.attr.struct.change}} about changing \verb+Deserialize+ to \verb+ConfigDeserialize+ is specific to that program, but other changes would apply more generally. Also, for \textit{\textbf{G.attr.struct.drop}}, we demonstrated a case where the developer removed a \verb+Debug+ attribute. A novel type of performance program repair tool working with a run-time profiler could identify unnecessary code and remove attributes that generate it.

@@

A struct enables users to define a custom type that is comprised of different types. This is a common construct across languages, and the patterns that we describe in this section also exist in other languages. \\

@@

\noindent\textbf{Description:} Adding \texttt{pub} to a struct field makes it possible to access that field from external modules. In the instances of this cluster, adding the \texttt{pub} access modifier happened in bug fixing changes where the developer needed to access a field which had not been marked for public access. On the other hand, public access might be revoked if the developer realizes there are no external accesses to the field, and that such accesses are not desirable, e.g. for encapsulation purposes. \\

\noindent\textbf{Actionability of struct changes:} The patterns related to access modifiers suggest an opportunity for repair tools to make sure that the accessibility of program entities follows the user's specifications, enhancing encapsulation, security, and maintainability. Additionally, the empirical numbers revealing the frequency of struct field modifications in Rust shed light on the significance of these bug fix patterns and emphasize the need for further exploration on how program repair tools can effectively modify code based on changes in struct fields to repair program behavior.

@@

\noindent\textbf{Description:} Rust is a null-safe language, meaning that object references cannot take on null values. If a developer wants a variable to contain the equivalent of null, they would have to use the \texttt{Option} type. \texttt{Option} is essentially an \texttt{enum} with two variants: \texttt{Some}, which indicates that a variable has a value; and \texttt{None}, which indicates the absence of any value. In this change pattern, the developer changes the type of a struct field to an \texttt{Option} of that type, hence enabling them to store no value in that field. 
For instance, in the Rust project swc-project/swc (a fast TypeScript/JavaScript compiler), the developer needed to modify the CSS parser source code to account for empty @media queries\footnote{\scriptsize \url{https://github.com/swc-project/swc/commit/75a14f98b7370226115ee24eec6eb8c802bd4837}}. As shown in the snippet above, they changed type \texttt{MediaQueryList} to \verb+Option<MediaQueryList>+, and modified the other parts of the source code accordingly. \\

\noindent\textbf{Actionability of adding Option to a type:} When the user's specification requires a variable to be able to assume \texttt{None} values, these patterns indicate the need to modify the program accordingly. A program repair\todo{is this really a repair tool or a refactoring tool? I feel like many of our changes are for refactorings not repairs, despite our PR} tool can leverage these patterns to automatically update the variable's type to \texttt{Option<T>}, ensuring that it can now hold \texttt{None} as a valid value. Furthermore, such a tool should also systematically update other occurrences of that variable throughout the program to account for the change and ensure program correctness.

@@

Rust is a statically-typed language, and like in other statically-typed languages, some change patterns are associated with changes in types used in different program structures. Here, we introduce four bug fix patterns that are associated with types in Rust:

@@

We collected 27143 datapoints for borrow-checker related fix patterns. We applied DBSCAN with parameters $\epsilon=0.001$ and $Z=5$, resulting in 102 clusters. After manual analysis (Section~\ref{sec:manual_analysis_cluster_selection}), we ended up with 8 borrow-checker related fix patterns. Table~\ref{table:bc} describes BC-related patterns, analogously to Table~\ref{table:general} for general patterns. Since the Borrow-Checker is unique to Rust, patterns in this section generally do not have direct counterparts in other languages. We next provide a detailed description of each of these patterns.

@@

\noindent\textbf{Actionability of clone removal:} Like Clippy, a program repair tool can implement clone removal by reasoning about whether the clone operation was necessary. If the clone is unnecessary, the tool can replace it with a reference (as seen above), thus reducing the overhead of unnecessary copying. \\

@@

\noindent\textbf{Actionability of adding borrowing:} This pattern often occurs as part of a larger change, making it challenging to create a generalized repair tool that covers all possible scenarios. Typically, when a reference is passed instead of transferring ownership, the variable is used again after returning from the call. Although it is not obvious how to create a comprehensive repair tool for such cases, empirical results underscore the significance and prevalence of this fix, warranting further investigation. \\

\noindent\textit{\textbf{BC.deref.to\_string().add} Adding a dereference before calling to\_string()}

@@

\noindent\textbf{Description:} This is a performance-related \todo{is it only perf?} change that in all instances (in our benchmarks) has been proposed by Clippy---commit messages indicated the use of Clippy for this fix. The change applies on an object prior to calling \verb+to_string()+ on that object. In this change, a dereference operator is added to that object. The reason behind the change is that the object is a reference type of \verb+T+, while \verb+T+ directly implements \verb+to_string()+. Adding the dereference makes the compiler use the specialized implementation of \verb+to_string()+ and not go through slower string formatting methods\footnote{\url{https://rust-lang.github.io/rust-clippy/master/index.html\#inefficient\_to\_string}}.

@@

\noindent\textbf{Actionability of adding mutability:} A repair tool can use this pattern to introduce mutability to the target variable and modify other parts of the program accordingly. For instance, if the target variable has associated \texttt{impl} blocks, the change may involve modifying method annotations to ensure that the calling object is mutable, aligning with the desired behaviour.\\
%specifications.\todo{how would the specs be provided? how could this be done as part of repair vs as part of refactoring?!} \\

@@

\noindent\textbf{Description:} The developer drops the mut keyword before a variable, as they do not mutate the variable. This redundant mutability is reported by Clippy and can also be removed by it. Such a change may help future-proof the code against unintended future changes.
